---
title: "Assng4"
author: "Saira Asif"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(rpart)
library(tree)
#install.packages("rpart.plot")
library(rpart.plot)
library(lars)
library(glmnet)
library(caret)
library(plotly)

```

## DATA

```{r}
# reading in training and testing dataset
tumor_train <- read_csv("/Users/sairaasif49/Downloads/trainset.csv")
tumor_test <- read_csv("/Users/sairaasif49/Downloads/testset.csv")

# No missing/NA values is any columns
apply(tumor_train, 2, function(x){sum(is.na(x))})

# No 0s in the predictor columns
apply(tumor_train, 2, function(x){sum(x==0)})

```

## elastic net (include Lasso and Ridge in your search grid) (Saira )

```{r}
class(tumor_train)
dim(tumor_train)

```



## KNN (Lisa)



## a fully grown classification/regression tree (CART) (Saira)

The training set is used to build a classification trees. 

```{r}

# Fitting a classification tree to tumor dataset
class_tree <- rpart(status ~ ., data=tumor_train, method = "class")

summary(class_tree)

# Visualizing the unpruned tree
rpart.plot(class_tree)


# Checking the order of variable importance
class_tree$variable.importance

# Checking model fit on testing dataset
pred_tree = predict(class_tree, tumor_test, type = "class")

# Compairing the preditced reuslt w
table_fit <- table(pred_tree,tumor_test$status)

TP = table_fit[1]
TN = table_fit[2,2]
FN = table_fit[2]
FP =  table_fit[1,2]

#Accuracy of the model is 90.58824
(Acc = (TP+TN)/(TP+TN+FP+FN)) * 100

#Sensitivity of model is 68.83117

(SEN = TP/(TP+TN)) * 100

# Specificity of the model is  96
(SPES = TN/(TN+FP)) * 100


```

We can also check if pruning the tree will improve results. Pruning selects the cp (complexity parameter) value associated with a shorter tree that minimizes the cross-validated error rate (xerror). 

```{r}
# Checking if pruning will improve reuslts 
#plotcp(fit.tree)
printcp(class_tree)

# Explicitly request the lowest cp value
bestcp <- class_tree$cptable[which.min(class_tree$cptable[,"xerror"]),"CP"]
pruned_tree <- prune(class_tree, cp = bestcp)
rpart.plot(pruned_tree)

# Alternate specification 
pred_prune = predict(pruned_tree, tumor_test, type="class")

pruned_table <- table(pred_prune, tumor_test$status)

TP2 = pruned_table[1]
TN2 = pruned_table[2,2]
FN2 = pruned_table[2]
FP2 =  pruned_table[1,2]

#Accuracy of the model is 90.58824
(Acc2 = (TP2+TN2)/(TP2+TN2+FP2+FN2)) * 100

#Sensitivity of model is 68.83117

(SEN2 = TP2/(TP2+TN2)) * 100

# Specificity of the model is  96
(SPES2 = TN2/(TN2+FP2)) * 100


```
There is no change in classification after pruning the tree. This suggests that pruning is not necessary for this tree


## support vector machine (Lisa)



## a bagged version of CART (Zee)


## random forests. (Zee)



## Final table 


