---
title: "Assng4"
author: "Saira Asif"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(rpart)
library(tree)
#install.packages("rpart.plot")
library(rpart.plot)
library(lars)
library(glmnet)
library(caret)
library(plotly)
library(randomForest)
library(ipred)
#install.packages("Metrics")
library(Metrics)
library(class)
library(e1071)
library(readr)
library(MASS)

```

## DATA

```{r}
# reading in training and testing dataset
#tumor_train <- read_csv("/Users/sairaasif49/Downloads/trainset.csv")
#tumor_test <- read_csv("/Users/sairaasif49/Downloads/testset.csv")

# No missing/NA values is any columns
apply(tumor_train, 2, function(x){sum(is.na(x))})

# No 0s in the predictor columns
apply(tumor_train, 2, function(x){sum(x==0)})

#convert status to factor
tumor_train$status<-as.factor(tumor_train$status)
tumor_test$status<-as.factor(tumor_test$status)
```

## elastic net (include Lasso and Ridge in your search grid) (Saira )

LASSO package glmnet requires the the response variable to be a vector and the set of predictor variables to be of the class data.matrix.

```{r}
#define response variable as vector 
status <- tumor_train$status
# dim is NULL as it is a vector, not a matrix
dim(status)

#define matrix of predictor variables
xs <- data.matrix(tumor_train[, -1])
dim(xs)
# Variance is different across predictors, so it has not been standardized, however glmnet automatically standardizes the variables
apply(xs, 2, var)

# ELASTIC NET

set.seed(12)

#efine the type of re-sampling as cross-validation using 10-fold method
enet <- trainControl(method = "cv", number = 10)

# Allows us to have more alpha values, from 0 to 1, to include both lasso and ridge regression in the analysis.
elasticGrid <- expand.grid(.alpha = seq(0, 1, length = 10), .lambda = seq(.5, 7.5, 1))
elasticGrid <- expand.grid(.alpha = seq(0, 1, length = 10), .lambda = seq(0, 1, length =10))

#cv.glmnet for computing penalized linear regression models.
def_elnet = train(status ~ ., data=tumor_train, method = "glmnet", trControl = enet, tuneGrid = elasticGrid)

# results for each combination of alpha and lambda
def_elnet$results

#lower alpha and lambda - better models
par(mfrow = c(1,2))
plot(def_elnet$results[ ,c(1,3)])
plot(def_elnet$results[ ,c(2,3)])

## highest possible Accuracy - best model
max_acc <- which.max(def_elnet$results[ ,3])
# The accuracy of this model is 94.38
def_elnet$results[max_acc,] 

# ridge regression seems the best for this dataset (alpha =0)
def_elnet$bestTune

# Compute the prediction error on the training set
train_enet <- glmnet(xs, status, type.measure = "class", family = "binomial", alpha = def_elnet$results[max_acc,1], lambda = def_elnet$results[max_acc,2])

train_enet

# Model coefficients
coef(def_elnet$finalModel, def_elnet$bestTune$lambda)

# Make predictions
predictions <- def_elnet %>% predict(tumor_test)

table(predictions, tumor_test[,1])

#perform k-fold cross-validation to find optimal lambda value that minimizes PE/MSE
cv_model <- cv.glmnet(xs, status, type.measure = "class", family = "binomial")
plot(cv_model)
coef(cv_model,s = c(cv_model$lambda.min,cv_model$lambda.1se))

prd.fl <- predict(cv_model, newx = data.matrix(tumor_test[,-1]), s = c(cv_model$lambda.min, cv_model$lambda.1se)) 

```

#Quadratic Discriminant Analysis (Lisa)

```{r}
# Import the data from CSV files
trainset <- read_csv("trainset.csv") # Read training data from trainset.csv file
testset <- read_csv("testset.csv") # Read test data from testset.csv file

# Fit QDA model using the training data
qda.fit = qda(status ~ . , data=trainset) # Status is the dependent variable and all other variables are independent.

# Print the fitted model
qda.fit

# Use the fitted model to predict the status of testset
qda.class = predict(qda.fit, testset)$class # Predict the class of the testset based on the fitted QDA model.

# Create a contingency table to compare predicted and actual status of testset
table(qda.class, testset$status)

# Calculate the accuracy of the prediction
accuracy = mean(qda.class==testset$status) # Calculate the accuracy of the predicted values based on the actual values in the testset.

# Print the accuracy
print(paste("Accuracy:", round(accuracy, 2))) # Print the accuracy of the predicted values.
```

## KNN (Lisa)

```{r}

# Import the data
trainset <- read_csv("trainset.csv")
testset <- read_csv("testset.csv")

# Separate the response variable from the predictors
train.X <- trainset[, 2:30]
train.Direction <- trainset[, 1]
test.X <- testset[, 2:30]
test.Direction <- testset[, 1]

# Set the seed for reproducibility
set.seed(123)

train_status <- trainset$status
test_status <- testset$status  

# Fit the model to knn() function with k=1 
knn.pred <- knn(train = train.X, test = test.X, cl = train_status, k = 1)

#Create a confusion matrix
confusion_matrix <- table(knn.pred, test_status)

# Print the confusion matrix
print(confusion_matrix)

# Calculate the accuracy of the model
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 2)))

# Fit the KNN model
knn.pred <- knn(train = train.X, test = test.X, cl = train_status, k = 10)

# Create a confusion matrix
confusion_matrix <- table(knn.pred, test_status)

# Print the confusion matrix
print(confusion_matrix)

# Calculate the accuracy of the model
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 2)))

```

The result using k=1 are pretty good with 92% of observations are correctly predicted. When improving the number of k to 10, the accuracy increased to 95%. 

## a fully grown classification/regression tree (CART) (Saira)

The training set is used to build a classification trees. 

```{r}

# Fitting a classification tree to tumor dataset
class_tree <- rpart(status ~ ., data=tumor_train, method = "class")

summary(class_tree)

# Visualizing the unpruned tree
rpart.plot(class_tree)


# Checking the order of variable importance
class_tree$variable.importance

# Checking model fit on testing dataset
pred_tree = predict(class_tree, tumor_test, type = "class")

# Compairing the preditced reuslt w
table_fit <- table(pred_tree,tumor_test$status)  ################# could also just do confusionMatrix(table(pred_tree,tumor_test$status)) to get stats on model

TP = table_fit[1]
TN = table_fit[2,2]
FN = table_fit[2]
FP =  table_fit[1,2]

#Accuracy of the model is 90.58824
(Acc = (TP+TN)/(TP+TN+FP+FN)) * 100

#Sensitivity of model is 88.33333
(SEN = TP/(TP+FN)) * 100

# Specificity of the model is  96
(SPES = TN/(TN+FP)) * 100


```

We can also check if pruning the tree will improve results. Pruning selects the cp (complexity parameter) value associated with a shorter tree that minimizes the cross-validated error rate (xerror). 

```{r}
# Checking if pruning will improve results 
#plotcp(fit.tree)
printcp(class_tree)

# Explicitly request the lowest cp value
bestcp <- class_tree$cptable[which.min(class_tree$cptable[,"xerror"]),"CP"]
pruned_tree <- prune(class_tree, cp = bestcp)
rpart.plot(pruned_tree)

# Alternate specification 
pred_prune = predict(pruned_tree, tumor_test, type="class")

pruned_table <- table(pred_prune, tumor_test$status)

TP2 = pruned_table[1]
TN2 = pruned_table[2,2]
FN2 = pruned_table[2]
FP2 =  pruned_table[1,2]

#Accuracy of the model is 90.58824
(Acc2 = (TP2+TN2)/(TP2+TN2+FP2+FN2)) * 100

#Sensitivity of model is 68.83117

(SEN2 = TP2/(TP2+FN2)) * 100

# Specificity of the model is  96
(SPES2 = TN2/(TN2+FP2)) * 100


```
There is no change in classification after pruning the tree. This suggests that pruning is not necessary for this tree


## support vector machine (Lisa)


```{r}
# Read the training and test sets from CSV files
trainset <- read_csv("trainset.csv")
testset <- read_csv("testset.csv")

# Extract the features (x) and target (y) variables from the training set
x <- trainset[,-1]
y <- trainset[,1]

# Remove any missing target values
y <- y[!is.na(y)]

# Combine x and y into a new data frame
dat <- data.frame(x = x, y=as.factor(y))

# Train a support vector machine (SVM) with a linear kernel on the training data
svmfit <- svm(y~., data=dat, kernel="linear", cost =1, scale=FALSE)

# Plot the data with the SVM decision boundary
plot(y~., data=dat, svmfit)

# Print a summary of the trained SVM model
summary(svmfit)

# Tune the SVM hyperparameters using cross-validation on the training data
svmfit <- svm(y~., data=dat, kernel='linear', cost=0.1, scale=FALSE)
set.seed(1)
tune.out <- tune(svm, y ~ ., data=dat, kernel='linear', ranges=list(cost = c(0.001, 0.01, 0.1, 1, 3,5,10,100)))

# Print a summary of the tuning results
summary(tune.out)

# Extract the best performing SVM model from the tuning results
bestmod <- tune.out$best.model
summary(bestmod)

# Extract the features (x) and target (y) variables from the test set
xtest <- testset[,-1]
ytest <- testset[,1]

# Remove any missing target values
ytest <- ytest[!is.na(ytest)]

# Combine xtest and ytest into a new data frame
testdat <- data.frame(x= xtest, y = as.factor(ytest))

# Use the best SVM model to predict the target variable for the test data
ypred <- predict(bestmod, testdat)

# Create a confusion matrix and calculate accuracy
conf_mat <- confusionMatrix(ypred, testdat$y)
accuracy <- conf_mat$overall["Accuracy"]

# Print the confusion matrix and accuracy
print(conf_mat$table)
cat("\nAccuracy:", round(accuracy, 2))
```

## a bagged version of CART (Zee)
```{r}
##bagging regression trees with 150 bootstrap replications
set.seed(123)
bag<-bagging(status~., data=tumor_train, nbagg=150, coob=T)
bag
#Out-of-bag estimate of misclassification error:  0.0455

#most import variables for model
imp <- data.frame(var=names(tumor_train[,-1]), imp=varImp(bag))

#sort variable importance descending
VI_plot <- imp[order(imp$Overall, decreasing=TRUE),]

#visualize variable importance with horizontal bar plot
barplot(VI_plot$Overall,
        names.arg=rownames(VI_plot),
        horiz=TRUE,
        col='steelblue',
        xlab='Variable Importance')

bag_CART_pred<-predict(bag, newdata=tumor_test, type="class")

#confusionmatrix
confusionMatrix(table(bag_CART_pred,tumor_test$status))
#Accuracy=92.94%
#Sensitivity=93.33%
#Specificity=92%
```



## random forests. (Zee)
```{r}
#for reproducible results
set.seed(751)

#Creating random forest model
randomforest_model<-randomForest(status~., data=tumor_train)
randomforest_model
#random forest classification with 500 trees generated at each node, which splits into 5 smaller nodes

#visualization of most important variables in the model
varImpPlot(randomforest_model) 

#testing the model on unseen data
randomforest_test <- predict(randomforest_model, newdata = tumor_test, type= "class")

#Confusion matrix
confusionMatrix(table(randomforest_test,tumor_test$status)) 
###Accuracy= 94%
##Sensitivity= 95%
##Specificity= 92%  
```

## Final table 

```{r}
tab <- matrix( ncol=3, nrow=6,byrow=TRUE)
rownames(tab) <- c("Elastic Net", "KNN", "CART", "SVM", "Bagged CART", "Random Forest")
colnames(tab) <- c("Acc", "Specificity", "Sensitivity")

tab[1,] <-
tab[3,] <- c(round(Acc*100, 2), round(SPES*100,2), round(SEN*100,2) )
tab[5,] <- c(92.94,92, 93.33)
tab[6,] <- c(94, 92, 95)
```

